### 2018-07-17 chapters/01-introduction/intro.tex
General structure (without 'background')

A typical workflow of a scientist who writes a scientific publication is to use a \gls*{wp} to write the publication and one or more \gls*{cas} for Verification, Analysis and Visualization, among other tasks. Especially in \gls*{stem}, \LaTeX\footnote{Note that technically \LaTeX{} is not a \gls*{wp} (\url{https://www.latex-project.org/about/}, seen 07/2017) like \textit{Microsoft Word}. However, since \LaTeX{} (an extension of \TeX) is the standard to write in \gls*{stem} and we only focus on mathematical writings in this thesis, we categorize \LaTeX{} as a \gls*{wp} as well.} has become the de facto standard for writing scientific publications over the last 30 years~\cites{LATEX:Standard}[559]{DigitalTypo}{Knuth}. \LaTeX{} enables printing of mathematical formulae in a structure similar to handwritten style. For example, consider a Jacobi polynomial~\cite[(18.3), table 1]{NIST:DLMF}
\begin{equation}\label{eq:P}
P_n^{(\alpha , \beta)}(\cos(a\Theta)).
\end{equation}
This formula is written in \LaTeX{} as
\begin{equation}\label{eq:P-tex}
\verb|P_n^{(\alpha,\beta)}(\cos(a\Theta))|.
\end{equation}

While \LaTeX{} focuses on displaying mathematics, \gls*{cas} focus on computations and user friendly syntax. Therefore, each system (such as \LaTeX{}, \Maple{} or \Mathematica) uses its own representation and syntax. Hence, a writer needs to constantly translate mathematical expressions from one representation to another and back again. Table~\ref{tab:JacobiP-usecase} shows four different representations for the same formula~(\ref{eq:P}).

@@

Such translations can become complex, if there are non-obvious differences between the used systems. For example, the \gls*{cas} could use a different order of the arguments, define a normalization of the function for better performances or even use different domains for the variables. Multivalued functions are particularly difficult~\cite{AISC:MultivaluedFunctions}. A \gls*{cas} usually defines so called \textit{branch cuts} to compute principal values of multivalued functions~\cite{Maple:Cuts}. Thereby, implementations of multivalued functions are discontinuous. In general, positioning branch cuts follow some conventions. They are mostly defined conveniently, such as a straight line. However, since a branch cut is a curve, it also can be defined as a spiral or it can be defined for a different position in the complex plane. The position of branch cuts can therefore varies from \gls*{cas} to \gls*{cas}~\cite{Branches:acot}. Figure~\ref{fig:acot-cut-compare} illustrates two examples of different branch cut positioning for the inverse trigonometric arccotangent function. While \Maple{} defines the branch cut at ${[-\infty\iunit, -\iunit]}$, ${[\iunit,\infty\iunit]}$ (figure~\ref{fig:acot-cut1}), \Mathematica{} defines the branch cut at ${[-\iunit, \iunit]}$ (figure~\ref{fig:acot-cut2}).

@@

Hence, a \gls*{cas} user needs to fully understand the properties and special definitions (such as the position of branch cuts) in the \gls*{cas} to avoid mistakes during a translation~\cite{Maple:Cuts}. In consequence, a manual translation process is not only laborious, but also prone to errors. Note that this general problem has been named to automatic \gls*{p2c} conversion~\cite{POM-Tagger}.

An automatic translation process is desirable. Providing translations for mathematical \LaTeX{} expressions is difficult, because the semantic information is absent. Semantics is the meaning of an expression. Consider the example of the Jacobi polynomial in~(\ref{eq:P}). A scientific reader might be able to conclude information like $P$ indicating the Jacobi polynomial and $n$ is being a non-negative integer. However, for a computer it is difficult to understand the expression~(\ref{eq:P-tex}). Since \LaTeX{} is the de facto standard for mathematical expressions, most \gls*{cas} provide import and export functions for \LaTeX{} expressions~\cite{Maple:ImportExport,Mathematica:ImportExport,Matlab:ImportExport,Sage:ImportExport}. However, because of the explained difficulties, those functions are very limited. In fact, they only work for expressions with unique semantic information, but fail for more complex expressions, such as our Jacobi polynomial example.

The \gls*{nist} has therefore developed a set of semantic \LaTeX{} macros~\cite{DLMF:Macros}. Each of these macros is tied with a unique well-defined mathematical object and linked to the corresponding definition in the \gls*{dlmf}~\cite{NIST:DLMF,NIST:DLMF:Paper,NIST:Handbook}. The semantic macros are used to semantically enhance the \gls*{dlmf}. An outgrowth of the \gls*{dlmf} project is the \gls*{drmf}~\cite{DRMF:14,DRMF:15}, which defines more semantic macros, especially for \gls*{opsf}. In table~\ref{tab:JacobiP-usecase} we have already seen an example of a semantic macro. We call \LaTeX{} expressions with semantic macros \textit{semantic \LaTeX{}} and without \textit{generic \LaTeX}, respectively. With semantic \LaTeX, it is possible to provide translations to \gls*{cas}.

Note that partial results of this Master's thesis have already been published at the 10$^\text{th}$ Conference on Intelligent Computer Mathematics 2017~\cite{CICM:Paper}. Techniques, approaches and results have been roughly introduced in the article and are discussed in more detail in this thesis. Pictures and ideas that have already been presented there are not explicitly referenced in the following.

Furthermore, note that the translator we will present in this thesis is based on not published code from the \gls*{mlp} project and the \gls*{pom}-tagger. We will also use the mentioned set of \Macro s, which are not published yet as well. Therefore, the current version of the translator is not online available yet. However, we are planning to publish the translator soon on the \gls*{drmf} website\footnote{\url{http://drmf.wmflabs.org}}.

Moreover, I will use 'we' rather than 'I' in the subsequent chapters of this thesis, since I published and discussed my ideas with others including my advisors Dr.~Moritz Schubotz and Professor Abdou Youssef, the hosts during research visits Howard S.~Cohl and J\"urgen Gerhard, and the students Joon Bang and Kevin Chen.\footnote{Special thanks to M.~Schubotz to give me the permission for using this phrase from his doctoral thesis \cite{MORITZ:Dis}.}

\section{Goals of the project}
In this thesis, we will present an automatic tool for translations between semantic \LaTeX{} and the \gls*{cas} \Maple{} and \Mathematica. The translation tool should be lite, in a sense that a user does not need a whole software package to use the program. It is also a part of the \gls*{drmf} project and should provide interactive translations in the \gls*{drmf}. Furthermore, it should provide additional information about the translation process and present solutions if an appropriate translation is not possible.

Another goal is the translator's extensibility in order to provide the basis for future work.

\section{Structure}
In the following chapters, we will discuss our translator and verification techniques. Prior to explaining implementation details, we need to introduce some basic background information in chapter~\ref{ch:background}. We start this chapter with an overview of related work (section~\ref{sec:related-work}). Multivalued functions and branch cuts, which were already mentioned, will be introduced in section~\ref{sec:math-background}. Besides the mathematical background, we will discuss semantics in \LaTeX{} in section~\ref{sec:semantics}. A brief introduction to grammar in languages and how we will use it to understand mathematical expressions will be given in section~\ref{sec:mlp}. The chapter is closed by some definitions for the following chapters.

The next chapter~\ref{ch:translator} is the main chapter of the thesis. We will explain implementations and discuss problems and solutions of and for the translator. It starts with an outline of the goals and our approaches in section~\ref{sec:gen-appr}. While we will present implementation details for the forward translations in section~\ref{sec:forward-translation}, we will present the developed backward translation in the following section~\ref{sec:backward-translation}.

In chapter~\ref{ch:evaluation} we will discuss, how we can verify or validate a translated expression. The chapter presents some approaches, such as round trip tests (section~\ref{sec:round-trip}) and numerical tests (section~\ref{sec:numerical-tests}). Extracted formulae from the \gls*{dlmf} and \gls*{drmf} created a comprehensive test suite for these approaches. Section~\ref{sec:test-summary} gives a summary of the results.

The results and the current status of the translation tool will be discussed in chapter~\ref{ch:results}. In conclusion, chapter~\ref{ch:conc-future-work} explains some approaches to further improve the explained techniques in future work.

>>>>

A typical workflow of a scientist who writes a scientific publication is to use a \gls*{wp} to write the paper and one or more \gls*{cas} for verification, analysis and visualization, among other tasks. Especially in \gls*{stem} literature, \LaTeX\footnote{Note that technically \LaTeX{} is not a \gls*{wp} (\url{https://www.latex-project.org/about/}, seen 07/2017) like \textit{Microsoft Word}. However, since \LaTeX{} (an extension of \TeX) is the standard to write articles in \gls*{stem} and we only focus on mathematical writings in this paper, we categorize \LaTeX{} as a \gls*{wp} as well.} has become the de facto standard for writing scientific publications over the past 30 years~\parencites{LATEX:Standard}[559]{DigitalTypo}{Knuth}. \LaTeX{} enables printing of mathematical formulae in a structure similar to handwritten style. For example, consider the Jacobi polynomial~\parencite[18.3 in table 1]{NIST:DLMF}
\begin{equation}\label{eq:P}
P_n^{(\alpha , \beta)}(\cos(a\Theta)),
\end{equation}
where $\alpha, \beta > -1$, and $n$ a nonnegative integer. 
This formula is written in \LaTeX{} as
\begin{equation}\label{eq:P-tex}
\verb|P_n^{(\alpha,\beta)}(\cos(a\Theta))|.
\end{equation}

While \LaTeX{} focuses on displaying mathematics, \gls*{cas} focus on computations and user friendly syntax. Therefore, each system uses different representations and syntax. Hence, a writer needs to constantly translate mathematical expressions from one representation to another and back again. Table~\ref{tab:JacobiP-usecase} shows four different representations for the same formula~(\ref{eq:P}).

@@

Such translations can become complex, if there are non-obvious distinctions between the used systems. For example, a \gls*{cas} could use a normalized form of the function or of their arguments for better performances. In this translation scenario, multivalued functions are particularly difficult~\parencite{AISC:MultivaluedFunctions}. A \gls*{cas} usually defines so called \textit{branch cuts} to compute principal values of multivalued functions~\parencite{Maple:Cuts}. Thereby, implementations of multivalued functions are discontinuous. In general, positioning branch cuts follow some conventions. They are mostly defined conveniently, such as a straight line. Technically, branch cut are curves and therefore can be defined in more complex shapes, such as spirals. The position of branch cuts can vary from \gls*{cas} to \gls*{cas}~\parencite{Branches:acot}. Figure~\ref{fig:acot-cut-compare} illustrates two examples of different branch cut positioning for the inverse trigonometric arccotangent function. While \Maple{} defines the branch cut at ${[-\infty\iunit, -\iunit]}$, ${[\iunit,\infty\iunit]}$ (figure~\ref{fig:acot-cut1}), \Mathematica{} defines the branch cut at ${[-\iunit, \iunit]}$ (figure~\ref{fig:acot-cut2}).

@@

Hence, a \gls*{cas} user needs to fully understand the properties and special definitions (such as the position of branch cuts) in the \gls*{cas} to avoid mistakes during a translation~\parencite{Maple:Cuts}. In consequence, a manual translation process is not only laborious, but also prone to errors. Note that this general problem has been named to automatic \gls*{p2c} conversion~\parencite{POM-Tagger}.

An automatic translation process is desirable. Providing translations for mathematical \LaTeX{} expressions is difficult, because the semantic information is absent. Semantics is the meaning of an expression. Consider the example of the Jacobi polynomial in~(\ref{eq:P}). A scientific reader might be able to conclude information like $P$ indicating the Jacobi polynomial and $n$ is being a non-negative integer. However, for a computer it is difficult to understand the expression~(\ref{eq:P-tex}). Since \LaTeX{} is the de facto standard for mathematical expressions, most \gls*{cas} provide import and export functions for \LaTeX{} expressions\footnote{Maple since version Maple 18, \textit{Produce output suitable for LaTeX 2e}, \url{http://www.maplesoft.com/support/help/Maple/view.aspx?path=latex} seen 06/2017;\\
Mathematica, \textit{Generating and Importing TeX}, \url{https://reference.wolfram.com/language/tutorial/GeneratingAndImportingTeX.html} seen 06/2017;\\
Matlab, \textit{\LaTeX{} form of symbolic expression in MATLAB}, \url{https://www.mathworks.com/help/symbolic/latex.html} seen 06/2017;\\
SageMath, \textit{Sage Math}, \url{http://doc.sagemath.org/html/en/tutorial/latex.html} seen 06/2017}. However, because of the explained difficulties, those functions are very limited. In fact, they only work for expressions with unique semantic information, but fail for more complex expressions, such as our Jacobi polynomial example.

The \gls*{nist} has therefore developed a set of semantic \LaTeX{} macros~\parencite{DLMF:Macros}. Each of these macros is tied with a unique well-defined mathematical object and linked to the corresponding definition in the \gls*{dlmf}~\parencite{NIST:DLMF,NIST:DLMF:Paper,NIST:Handbook}. The semantic macros are used to semantically enhance the \gls*{dlmf}. An outgrowth of the \gls*{dlmf} project is the \gls*{drmf}~\parencite{DRMF:14,DRMF:15}, which defines more semantic macros, especially for \gls*{opsf}. In table~\ref{tab:JacobiP-usecase} we have already seen an example of a semantic macro. We call \LaTeX{} expressions with semantic macros \textit{semantic \LaTeX{}} and without \textit{generic \LaTeX}, respectively.

Note that partial results of this paper have been published in~\parencite{CICM:Paper}.

%Furthermore, note that the translator we will present in this thesis is based on not published code from the \gls*{mlp} project and the \gls*{pom}-tagger. We will also use the mentioned set of \Macro s, which are not published yet as well. Therefore, the current version of the translator is not online available yet. However, we are planning to publish the translator soon on the \gls*{drmf} website\footnote{\url{http://drmf.wmflabs.org}}.

%Moreover, I will use 'we' rather than 'I' in the subsequent chapters of this thesis, since I published and discussed my ideas with others including my advisors Dr.~Moritz Schubotz and Professor Abdou Youssef, the hosts during research visits Howard S.~Cohl and J\"urgen Gerhard, and the students Joon Bang and Kevin Chen.\footnote{Special thanks to M.~Schubotz to give me the permission for using this phrase from his doctoral thesis \parencite{MORITZ:Dis}.}

%\section{Goals of the project}
%In this thesis, we will present an automatic tool for translations between semantic \LaTeX{} and the \gls*{cas} \Maple{} and \Mathematica. The translation tool should be lite, in a sense that a user does not need a whole software package to use the program. It is also a part of the \gls*{drmf} project and should provide interactive translations in the \gls*{drmf}. Furthermore, it should provide additional information about the translation process and present solutions if an appropriate translation is not possible.

%Another goal is the translator's extensibility in order to provide the basis for future work.

%\section{Structure}
%In the following chapters, we will discuss our translator and verification techniques. Prior to explaining implementation details, we need to introduce some basic background information in chapter~\ref{ch:background}. We start this chapter with an overview of related work (section~\ref{sec:related-work}). Multivalued functions and branch cuts, which were already mentioned, will be introduced in section~\ref{sec:math-background}. Besides the mathematical background, we will discuss semantics in \LaTeX{} in section~\ref{sec:semantics}. A brief introduction to grammar in languages and how we will use it to understand mathematical expressions will be given in section~\ref{sec:mlp}. The chapter is closed by some definitions for the following chapters.

%The next chapter~\ref{ch:translator} is the main chapter of the thesis. We will explain implementations and discuss problems and solutions of and for the translator. It starts with an outline of the goals and our approaches in section~\ref{sec:gen-appr}. While we will present implementation details for the forward translations in section~\ref{sec:forward-translation}, we will present the developed backward translation in the following section~\ref{sec:backward-translation}.

%In chapter~\ref{ch:evaluation} we will discuss, how we can verify or validate a translated expression. The chapter presents some approaches, such as round trip tests (section~\ref{sec:round-trip}) and numerical tests (section~\ref{sec:numerical-tests}). Extracted formulae from the \gls*{dlmf} and \gls*{drmf} created a comprehensive test suite for these approaches. Section~\ref{sec:test-summary} gives a summary of the results.

%The results and the current status of the translation tool will be discussed in chapter~\ref{ch:results}. In conclusion, chapter~\ref{ch:conc-future-work} explains some approaches to further improve the explained techniques in future work.

### 2018-07-17 chapters/02-background/math-background.tex
General structure (without 'background')

A three-dimensional plot of a complex function, such as $f$ from above, usually maps the real and imaginary part of the variable to the $x$ and $y$ axes, while the height is defined by the absolute value of the complex variable $|z|$. Figure~\ref{fig:cylinderU} demonstrates a complex plot of the parabolic cylinder function $\ParabolicU@{a}{z}$~\cite[(12.2i)]{NIST:DLMF} in \Maple.

\begin{figure}[ht]
	\centering
	\includegraphics[width=0.8\textwidth]{cylinderUPlot.png}
	\caption{A 3-dimensional complex plot of the parabolic cylinder function $\ParabolicU@{a}{z}$ at $a=0$ plotted by \Maple.~\cite[(12.2i)]{NIST:DLMF}}
	\label{fig:cylinderU}
\end{figure}

@@

\begin{figure}[!htp]
    \centering
    \subfloat[Plot of the real value of $\ParabolicU@{0}{z}$.]{%
        \includegraphics[width=0.8\textwidth]{cylinderURe.png}
        \label{fig:cylinderURe}%
    }
    \vspace{-0.1cm}
    \subfloat[Plot of the imaginary value of $\ParabolicU@{0}{z}$.]{%
        \includegraphics[width=0.8\textwidth]{cylinderUIm.png}
        \label{fig:cylinderUIm}%
    }
    \caption{Separately plot of the real~\protect\subref{fig:cylinderURe} and imaginary~\protect\subref{fig:cylinderUIm} part of the parabolic cylinder function $\ParabolicU@{a}{z}$ at $a=0$.}
    \label{fig:cylinderUSplit}
\end{figure}

@@

A more complicated example might be the equation of the already introduced parabolic cylinder function $\ParabolicU@{a}{z}$ and the modified Bessel function of the second kind $\BesselK{\nu}@{z}$. The parabolic cylinder function is a solution of a second order differential equation~\cite[(12.2i)]{NIST:DLMF} and can be represented by $\BesselK{\nu}@@{z}$ for $a=0$. The relation is defined in~(\ref{eq:branch-cut-issue}).

\begin{equation}\label{eq:branch-cut-issue}
\displaystyle \ParabolicU@{0}{z} = \sqrt{\frac{z}{2\cpi}}\BesselK{\frac{1}{4}}@{\frac{1}{4}z^2}
\end{equation}

>>>>

A three-dimensional plot of a complex function, such as $f$ from above, usually maps the real and imaginary part of the variable to the $x$ and $y$ axes, while the height is defined by the absolute value of the complex variable $|z|$. Figure~\ref{fig:cylinderU} demonstrates a complex plot of the parabolic cylinder function $U(a,z)$~\cite[(12.2i)]{NIST:DLMF} in \Maple.

\begin{figure}[ht]
	\centering
	\includegraphics[width=0.8\textwidth]{cylinderUPlot.png}
	\caption{A 3-dimensional complex plot of the parabolic cylinder function $U(a,z)$ at $a=0$ plotted by \Maple.~\cite[(12.2i)]{NIST:DLMF}}
	\label{fig:cylinderU}
\end{figure}

@@

\begin{figure}[!htp]
    \centering
    \subfloat[Plot of the real value of $\ParabolicU@{0}{z}$.]{%
        \includegraphics[width=0.8\textwidth]{cylinderURe.png}
        \label{fig:cylinderURe}%
    }
    \vspace{-0.1cm}
    \subfloat[Plot of the imaginary value of $\ParabolicU@{0}{z}$.]{%
        \includegraphics[width=0.8\textwidth]{cylinderUIm.png}
        \label{fig:cylinderUIm}%
    }
    \caption{Separately plot of the real~\protect\subref{fig:cylinderURe} and imaginary~\protect\subref{fig:cylinderUIm} part of the parabolic cylinder function $U(a,z)$ at $a=0$.}
    \label{fig:cylinderUSplit}
\end{figure}

@@

A more complicated example might be the equation of the already introduced parabolic cylinder function $U(a,z)$ and the modified Bessel function of the second kind $\BesselK{\nu}@{z}$. The parabolic cylinder function is a solution of a second order differential equation~\cite[(12.2i)]{NIST:DLMF} and can be represented by $\BesselK{\nu}@@{z}$ for $a=0$. The relation is defined in~(\ref{eq:branch-cut-issue}).

\begin{equation}\label{eq:branch-cut-issue}
\displaystyle U(0,z) = \sqrt{\frac{z}{2\cpi}}\BesselK{\frac{1}{4}}@{\frac{1}{4}z^2}
\end{equation}

### 2018-07-17 chapters/02-background/related-work.tex
General structure (without 'background')

\section{Related Work}\label{sec:related-work}
Since \LaTeX{} became the de facto standard for writing papers in mathematics, most of the \gls{cas} nowadays are able to import and export mathematical \LaTeX{} expressions~\cite{Maple:ImportExport,Mathematica:ImportExport,Matlab:ImportExport,Sage:ImportExport}. Those tools have two essential problems. First of all, they cannot import mathematical expressions where the semantic information is absent, which is the usual case for generic \LaTeX{} expressions. Therefore, they are only able to import simple mathematical expressions, where the semantics are unique. For example, the internal \LaTeX{} macro \verb|\frac| always indicates a fraction. These import tools fail for more complex expressions, for example the Jacobi polynomial in \cref{tab:JacobiP-usecase}. The second problem appears in the export tools. Mathematical expressions in \gls{cas} are fully semantic, otherwise the \gls{cas} wouldn't be able to compute or evaluate the expressions. During the export process, the semantic information gets lost, because generic \LaTeX{} is not able to carry semantic information. In consequence of these two problems, an exported expression cannot be imported to the same system again in most cases (except those simple expressions described above). Our tool should solve these problems and provide round trip translations between \LaTeX{} and \gls{cas}.

@@

However, the pre-definition approach is simple and easy to realize, as long as the system can carry semantic information. A typical, not mathematical example are Wikipedia articles~\cite{Wiki}. An author of such an article can specify further information about symbols, words and sentences by creating a hyperlink to a more detailed explanation. Usually, the author defines the hyperlink, when he is writing the article or he needs to manually add the hyperlink later on. Similar to this approach are \textit{interactive documents}\footnote{There is no adequate definition what interactive documents are. However, this name is widely used to describe electronic document formats that allow interactivity to change the content in real time.}, such as the \gls{cdf}~\cite{CDF:Wolfram} by Wolfram Research or the \textit{worksheets} of \Maple{}, which are also sometimes referred to as interactive documents. Those documents pre-define the semantics to allow computations. Those complex document formats require specialized tools to show and work with the documents (Wolfram CDF Player, or \Maple{} for the \textit{worksheets}). The \JOBAD{} architecture~\cite{JOBAD:orig} is able to create web based interactive documents and uses \gls{omdoc}~\cite{OMDoc} to carry semantics. The documents can be viewed and edited in the browser. Those \JOBAD{-documents} also allow to perform computations during \gls{cas}. This gives the opportunity to calculate, compute and change mathematical expressions directly in the document. The translation performs in the background, invisible for the user. Similar to the \JOBAD{} architecture other interactive web documents exist, such as \textit{MathDox}~\cite{MathDox} and \textit{The Planetary System}~\cite{Planetary}. All of them demonstrate the potential for the education system.

All of these systems carry semantic information. Most of them create their own way to make this possible. The web based documents mostly use standardized ways, such as the content \gls{mathML}, a specialization of \gls{mathML}~\cite{MathML}, that focuses on the semantics of an expression rather than any particular rendering for the expression. Since \LaTeX{} is the de facto standard to write mathematical expressions in documents there are projects to extend \LaTeX{} in a way that enable to carry the semantic information directly in \LaTeX. There are two known projects that created a semantic version of \LaTeX: \sTeX{}~\cite{sTeX} developed by Michael Kohlhase and the \Macro s developed by Bruce Miller. Our translator uses the \Macro s rather than \sTeX. A full explanation of this decision will be given in \cref{sec:semantics}.

@@

Another approach that tries to avoid the translation problem are \LaTeX{} packages that allow computations directly through the \LaTeX{} compiler (for example \textit{LaTeXCalc}~\cite{LatexCalc}) or allow \gls{cas} commands for computations directly in \TeX -files. For example, \textit{sagetex}~\cite{Sagetex} is an interface \LaTeX{} package for the open source \gls{cas} \textit{sage}\footnote{An abbreviation for \textit{SageMath}.}. This package allows \textit{sage} commands in \TeX -files and uses \textit{sage} in the background to compute the commands. Obviously, the first example is not as powerful as a \gls{cas} would be. However, \textit{sagetex} does not really solve our problem of the scientific workflow. The writer still needs to translate expressions manually. The only difference is that the author can write computations directly into the \TeX -file. However, these approaches have the potential for a powerful combination with our translation tool.

>>>>

\section{Related Work}\label{sec:related-work}
Since \LaTeX{} became the de facto standard for writing papers in mathematics, most of the \gls{cas} nowadays are able to import and export mathematical \LaTeX{} expressions. Those tools have two essential problems. First of all, they cannot import mathematical expressions where the semantic information is absent, which is the usual case for generic \LaTeX{} expressions. Therefore, they are only able to import simple mathematical expressions, where the semantics are unique. For example, the internal \LaTeX{} macro \verb|\frac| always indicates a fraction. These import tools fail for more complex expressions, for example the Jacobi polynomial in \cref{tab:JacobiP-usecase}. The second problem appears in the export tools. Mathematical expressions in \gls{cas} are fully semantic, otherwise the \gls{cas} wouldn't be able to compute or evaluate the expressions. During the export process, the semantic information gets lost, because generic \LaTeX{} is not able to carry semantic information. In consequence of these two problems, an exported expression cannot be imported to the same system again in most cases (except those simple expressions described above). Our tool should solve these problems and provide round trip translations between \LaTeX{} and \gls{cas}.

@@

However, the pre-definition approach is simple and easy to realize, as long as the system can carry semantic information. A typical, not mathematical example are Wikipedia articles~\parencite{Wiki}. An author of such an article can specify further information about symbols, words and sentences by creating a hyperlink to a more detailed explanation. Usually, the author defines the hyperlink, when he is writing the article or he needs to manually add the hyperlink later on. Similar to this approach are \textit{interactive documents}\footnote{There is no adequate definition what interactive documents are. However, this name is widely used to describe electronic document formats that allow interactivity to change the content in real time.}, such as the \gls{cdf}~\parencite{CDF:Wolfram} by Wolfram Research or the \textit{worksheets} of \Maple{}, which are also sometimes referred to as interactive documents. Those documents pre-define the semantics to allow computations. Those complex document formats require specialized tools to show and work with the documents (Wolfram CDF Player, or \Maple{} for the \textit{worksheets}). The \JOBAD{} architecture~\parencite{JOBAD:orig} is able to create web based interactive documents and uses \gls{omdoc}~\parencite{OMDoc} to carry semantics. The documents can be viewed and edited in the browser. Those \JOBAD{-documents} also allow to perform computations during \gls{cas}. This gives the opportunity to calculate, compute and change mathematical expressions directly in the document. The translation performs in the background, invisible for the user. Similar to the \JOBAD{} architecture other interactive web documents exist, such as \textit{MathDox}~\parencite{MathDox} and \textit{The Planetary System}~\parencite{Planetary}. All of them demonstrate the potential for the education system.

All of these systems carry semantic information. Most of them create their own way to make this possible. The web based documents mostly use standardized ways, such as the content \gls{mathML}, a specialization of \gls{mathML}~\parencite{MathML}, that focuses on the semantics of an expression rather than any particular rendering for the expression. Since \LaTeX{} is the de facto standard to write mathematical expressions in documents there are projects to extend \LaTeX{} in a way that enable to carry the semantic information directly in \LaTeX. There are two known projects that created a semantic version of \LaTeX: \sTeX{}~\parencite{sTeX} developed by Michael Kohlhase and the \Macro s developed by Bruce Miller. Our translator uses the \Macro s rather than \sTeX. A full explanation of this decision will be given in \cref{sec:semantics}.

@@

Another approach that tries to avoid the translation problem are \LaTeX{} packages that allow computations directly through the \LaTeX{} compiler (for example \textit{LaTeXCalc}~\parencite{LatexCalc}) or allow \gls{cas} commands for computations directly in \TeX -files. For example, \textit{sagetex}~\parencite{Sagetex} is an interface \LaTeX{} package for the open source \gls{cas} \textit{sage}\footnote{An abbreviation for \textit{SageMath}.}. This package allows \textit{sage} commands in \TeX -files and uses \textit{sage} in the background to compute the commands. Obviously, the first example is not as powerful as a \gls{cas} would be. However, \textit{sagetex} does not really solve our problem of the scientific workflow. The writer still needs to translate expressions manually. The only difference is that the author can write computations directly into the \TeX -file. However, these approaches have the potential for a powerful combination with our translation tool.

### 2018-07-17 chapters/03-translator/backward-trans.tex
General structure (without 'background')

Note that all explanations in this section is based on \Maple's programming guide~\cite{MAPLE:ProgrammingGuide} for \Maple{} 2016.

@@

\begin{table}[ht]
\centering
\begin{tabular}{cp{10cm}}
	\hline
	Type & Explanation\\
	\hline
	SUM & Sums.\\
	PROD & Products.\\
	EXPSEQ & Expression sequence is a kind of list. The arguments of functions are stored in such sequences.\\
	INTPOS & Positive integers.\\
	INTNEG & Negative integers.\\
	COMPLEX & Complex numbers with real and imaginary part.\\
	FLOAT & Float numbers are stored in the scientific notation with integer values for the exponent $n$ and the significand $m$ in $m \cdot 10^n$.\\
	RATIONAL & Rational numbers are fractions stored in integer values for the numerator and positive integers for the denominator.\\
	POWER & Exponentiation with expressions as base and exponent.\\
	FUNCTION & Function invocation with the name, arguments and attributes of the function.\\
	\hline
\end{tabular}
\caption{A subset of important internal \Maple{} structures. See~\cite{MAPLE:ProgrammingGuide} for a complete list.}
\label{tab:maple-types}
\end{table}

\subsection{Maple's Open Maple API}
\Maple{} provides an \gls{api}, called \texttt{OpenMaple}~\cite[\S 14.3]{MAPLE:ProgrammingGuide} for interaction with the \Maple{} kernel. The \texttt{OpenMaple} \gls{api} is implemented for different programming languages such as \texttt{C}, \texttt{Java} and \texttt{Visual Basic}. Since our forward translation and the \gls{mlp} is implemented in \texttt{Java}, we use the \texttt{Java} \gls{api} for the backward translation. The \texttt{Java} \gls{api} has some limitations to interact with \Maple's \gls{dag}.

@@

The following figure~\ref{fig:backward-trans} illustrates the backward translation process for the Jacobi polynomial example $\JacobiP{\alpha}{\beta}{n}@{\cos@{a\Theta}}$ from table~\ref{tab:JacobiP-usecase}. The input expression is converted to the nested list representation of the \texttt{MyInertForm} (the customized \inertF{} of \Maple). Afterwards, each node of the tree is translated separately by specialized subtranslators (visualized by blue arrows). The translation of functions (bold blue arrows) is again realized by translation patterns, which define the correct position for each argument. The last step is used to put all translated arguments to the right position (visualized by red arrows).

\begin{figure}[!htp]
	\centering
	\includegraphics[clip, trim=0.1cm 0.1cm 0.1cm 0.1cm, scale=0.7]{MapleTranslation.pdf}
	\caption{A scheme of the backward translation process from \Maple{} for the Jacobi polynomial $\JacobiP{\alpha}{\beta}{n}@{\cos@{a\Theta}}$. The input string is converted by the \Maple{} kernel into the nested list representation. This list is translated by subtranslators (blue and red arrows). A function translation (bold blue arrows) is again realized by translation patterns to define the position of the arguments (red arrows).}
	\label{fig:backward-trans}
\end{figure}

>>>>

Note that all explanations in this section is based on \Maple's programming guide~\parencite{MAPLE:ProgrammingGuide} for \Maple{} 2016.

@@

\begin{table}[ht]
\centering
\begin{tabular}{cp{10cm}}
	\hline
	Type & Explanation\\
	\hline
	SUM & Sums.\\
	PROD & Products.\\
	EXPSEQ & Expression sequence is a kind of list. The arguments of functions are stored in such sequences.\\
	INTPOS & Positive integers.\\
	INTNEG & Negative integers.\\
	COMPLEX & Complex numbers with real and imaginary part.\\
	FLOAT & Float numbers are stored in the scientific notation with integer values for the exponent $n$ and the significand $m$ in $m \cdot 10^n$.\\
	RATIONAL & Rational numbers are fractions stored in integer values for the numerator and positive integers for the denominator.\\
	POWER & Exponentiation with expressions as base and exponent.\\
	FUNCTION & Function invocation with the name, arguments and attributes of the function.\\
	\hline
\end{tabular}
\caption{A subset of important internal \Maple{} structures. See~\parencite{MAPLE:ProgrammingGuide} for a complete list.}
\label{tab:maple-types}
\end{table}

\subsection{Maple's Open Maple API}
\Maple{} provides an \gls{api}, called \texttt{OpenMaple}~\parencite[\S 14.3]{MAPLE:ProgrammingGuide} for interaction with the \Maple{} kernel. The \texttt{OpenMaple} \gls{api} is implemented for different programming languages such as \texttt{C}, \texttt{Java} and \texttt{Visual Basic}. Since our forward translation and the \gls{mlp} is implemented in \texttt{Java}, we use the \texttt{Java} \gls{api} for the backward translation. The \texttt{Java} \gls{api} has some limitations to interact with \Maple's \gls{dag}.

@@

The following figure~\ref{fig:backward-trans} illustrates the backward translation process for the Jacobi polynomial example $P_n^{(\alpha , \beta)}(\cos(a\Theta))$ from table~\ref{tab:JacobiP-usecase}. The input expression is converted to the nested list representation of the \texttt{MyInertForm} (the customized \inertF{} of \Maple). Afterwards, each node of the tree is translated separately by specialized subtranslators (visualized by blue arrows). The translation of functions (bold blue arrows) is again realized by translation patterns, which define the correct position for each argument. The last step is used to put all translated arguments to the right position (visualized by red arrows).

\begin{figure}[!htp]
	\centering
	\includegraphics[clip, trim=0.1cm 0.1cm 0.1cm 0.1cm, scale=0.7]{MapleTranslation.pdf}
	\caption{A scheme of the backward translation process from \Maple{} for the Jacobi polynomial $P_n^{(\alpha , \beta)}(\cos(a\Theta))$. The input string is converted by the \Maple{} kernel into the nested list representation. This list is translated by subtranslators (blue and red arrows). A function translation (bold blue arrows) is again realized by translation patterns to define the position of the arguments (red arrows).}
	\label{fig:backward-trans}
\end{figure}
